---
title: "Customer Segmentation Analysis"
author: "Ahomagnon Romuald"
date: "2025-02-05"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(cluster)
library(factoextra)
library(NbClust)
library(plotrix)
library(plotly)
library(GGally)
library(dbscan)
```
# 1. Data Loading and Preprocessing

```{r}
# Load the dataset
Data_custumer <- read.csv("~/Downloads/Mes projets/customer-segmentation-dataset/Mall_Customers.csv")

# Normalize the data (scale Age, Annual Income, and Spending Score)
Data_custumer[, 3:5] <- scale(Data_custumer[, 3:5])

# Display the first few rows of the dataset
head(Data_custumer)
```
2. Data Exploration and Visualization
Pairwise Relationships

```{r}
# Visualize pairwise relationships between variables
ggpairs(Data_custumer[, 3:5], title = "Pairwise Relationships Between Variables")
```

# Gender Distribution

```{r}
# Bar plot for gender distribution
a <- table(Data_custumer$Gender)
barplot(a, main = "Gender Comparison", ylab = "Count", xlab = "Gender", col = rainbow(2), legend = rownames(a))

# 3D Pie Chart for gender distribution
pct <- round(a/sum(a)*100)
lbs <- paste(c("Female", "Male"), " ", pct, "%", sep = " ")
pie3D(a, labels = lbs, main = "Ratio of Female and Male Customers")
```



Age Distribution
```{r}
# Histogram for age distribution
hist(Data_custumer$Age, col = "blue", main = "Age Distribution", xlab = "Age Class", ylab = "Frequency", labels = TRUE)

# Boxplot for age distribution
boxplot(Data_custumer$Age, col = "#ff0066", main = "Boxplot for Age Distribution")

```



Annual Income Analysis
```{r}

# Histogram for annual income
hist(Data_custumer$Annual.Income..k.., col = "#660033", main = "Annual Income Distribution", xlab = "Annual Income (k$)", ylab = "Frequency", labels = TRUE)

# Density plot for annual income
plot(density(Data_custumer$Annual.Income..k..), col = "yellow", main = "Density Plot for Annual Income", xlab = "Annual Income (k$)", ylab = "Density")
polygon(density(Data_custumer$Annual.Income..k..), col = "#ccff66")

```


Spending Score Analysis
```{r}

# Boxplot for spending score
boxplot(Data_custumer$Spending.Score..1.100., horizontal = TRUE, col = "#990000", main = "Boxplot for Spending Score")

# Histogram for spending score
hist(Data_custumer$Spending.Score..1.100., main = "Spending Score Distribution", xlab = "Spending Score", ylab = "Frequency", col = "#6600cc", labels = TRUE)

```


3. Clustering Analysis
Elbow Method
```{r}

# Function to calculate total intra-cluster sum of squares (WCSS)
iss <- function(k) {
  kmeans(Data_custumer[, 3:5], k, iter.max = 100, nstart = 100, algorithm = "Lloyd")$tot.withinss
}

# Compute WCSS for k = 1 to 10
k.values <- 1:10
iss_values <- sapply(k.values, iss)

# Plot the elbow curve
plot(k.values, iss_values, type = "b", pch = 19, frame = FALSE, xlab = "Number of Clusters (k)", ylab = "Total Intra-Cluster Sum of Squares", main = "Elbow Method")

```

Silhouette Method

```{r}
# Silhouette Method
fviz_nbclust(Data_custumer[, 3:5], kmeans, method = "silhouette") +
  ggtitle("Silhouette Method for Optimal Number of Clusters")
```

Gap Statistic Method

```{r}
# Gap Statistic Method
set.seed(125)
stat_gap <- clusGap(Data_custumer[, 3:5], FUN = kmeans, nstart = 25, K.max = 10, B = 50)
fviz_gap_stat(stat_gap) +
  ggtitle("Gap Statistic Method for Optimal Number of Clusters")


```



Final K-means Clustering
```{r}
# Perform K-means clustering with the optimal number of clusters (k = 6)
k6 <- kmeans(Data_custumer[, 3:5], centers = 6, iter.max = 100, nstart = 50, algorithm = "Lloyd")
k6
```



4. Cluster Validation
Davies-Bouldin Index
```{r}
# Install and load the clusterSim package
install.packages("clusterSim")
library(clusterSim)

# Compute Davies-Bouldin Index for cluster validation
davies_bouldin <- index.DB(Data_custumer[, 3:5], k6$cluster)
davies_bouldin
```


```{r}
# Create a 3D scatter plot of clusters
plot_ly(Data_custumer, x = ~Annual.Income..k.., y = ~Spending.Score..1.100., z = ~Age, color = ~as.factor(k6$cluster), colors = c("#FF5733", "#33FF57", "#3357FF", "#FF33A1", "#A133FF", "#33FFF5")) %>%
  layout(title = "3D Scatter Plot of Customer Segments")

```



Cluster Profiles
```{r}

# Summarize cluster profiles
cluster_profiles <- aggregate(Data_custumer[, 3:5], by = list(k6$cluster), FUN = mean)
cluster_profiles

```


6. Conclusion

    Key Findings: Identified 6 distinct customer segments based on income, spending score, and age.

    Business Implications: Tailor marketing strategies to target each segment effectively.



